---
title: "Cluster Analysis"
author: "Fin Mooney"
date: "2023-05-09"
output: html_document
---

## Resources

### Intro to Spatial Data in R:

https://cengel.github.io/R-spatial/intro.html

### Point Pattern Analysis Example:

http://rstudio-pubs-static.s3.amazonaws.com/481933_61ed3eff92e74c04a52d44a48b1b4548.html

### Intro to Spatial Autocrrelation Analysis in R
https://rpubs.com/quarcs-lab/spatial-autocorrelation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(sp)
require(sf)
require(spdep)
require(spatstat)
require(rgdal)
```








## Load Vector Spatial Data

```{r load vector spatial data, eval=FALSE, echo=TRUE}
## The sp package is used to download shapefiles 

#Insert the folder location that has the shapefile into st_read
sf <- st_read("folder_path")

## If you only have a tabular file with coordinates, you need to convert it into a sf object 

# Create coords vector that specifies that data frame's columns for longitude and latitude

df <- read.csv("file_path")
sf <- st_as_sf(df, coords = c("POINT_X", "POINT Y"))

## Attach a Coordinate System to the SF
st_crs(sf) # Will come out as NA

st_crs(sf) <- 4326 #insert the EPSF code for the projection, in this example was are using the code for WGS84

## Save the created shapefile to a location on your hard drive

st_write(sf, "folder_location", driver = "ESRI Shapefile")
```

``` {r spatialdataframe, eval=FALSE, echo=TRUE}

## If we have a dataframe with values attached to the points we can create a SpatialPointsDataFrame object using the sp package

## Need to specify coordinates
coordinates(df) <- c("POINT_X", "POINT_Y")
class(df)

## Add spatial reference

# The rgdal package is needed to save the sp object as a shapefile with a coordinate system 

proj4string(df) <- CRS("+init=epsg:4326") # this is WGS84
is.projected(df) 

```

## Visualize SF Object

``` {r visualize sf object, eval=FALSE, echo=TRUE}

# To vilsualize SF object we can use ggplot2 for basic mapping or the leaflet package for web mapping

ggplot(sf, aes(name_of_column_with_x_value, name_of_column_with_y_value)) +
  geom_point()

# OR

ggplot(sf) + 
  geom_sf(aes(fill=vaule_column)) # sf must be sf_object

```

## K and L Plots

- K Plot = Summarizies the distance between points for all distances
- K values above K expected: clustering of points at given distance band
- K Values below K expected: dispersion of points at a given distance band


- L Plot = Helps determine small differences between K and K-expected
- L values above zero : clustering of points at a given distance band
- L values below zero : dispersion of points at a given distance band (i.e., inhibition)

## Setting Up Spatstat Object

```{r Spatstat, eval=FALSE, echo=TRUE}
## The spatstat package requires that your point data be transformed into ppp object

# need to create a window for the boundary of the point
win <- as.owin(sf_boundary) # insert boundary (e.g, census tract, country) of the points, note: boundary must be a sf object

class(win) #check it worked

# Get the coordinates of the point data
coords <- st_coordinates(sf_points)
head(coords)

p <- ppp(coords[,1], coords[,2], window=win)
par(mai=c(0,0,0,0))
plot(p) # plot the ppp object
```

## K-plot

```{r Kplot, eval=FALSE, echo=TRUE}
# Calculate the K-function
K <- Kest(p)

# Plot the K-function
plot(K, main = "K-function for Points")
```

## L-Plot

``` {r lplot, eval=FALSE, echo=TRUE}
# Calculate the L-function
 L <- Lest(p)

# Plot the L-Function
 plot(L, main="L-Function for Points")

```

## GLobal Moran's I 
- Measure of spatial autocorrelation 
- Compare clustering of points across areas
- Ranges from -1 (indicating perfect negative spatial autocorrelation) to 1 (indicating perfect positive spatial autocorrelation), with 0 indicating no spatial autocorrelation. 
- Significant positive Moran's I indicates that high values of the variable tend to cluster together, while a significant negative Moran's I indicates that low values of the variable tend to cluster together.

``` {r GlobalMoransI, eval=FALSE, echo=TRUE}
# Need coordiantes of your points
coords <- st_coordinates(points)

# Create a spatial weights matrix, when conducting a GLobal Moran's I on points we will need to use distance-based weights

dists <- dnearneigh(coords, 0, 100) # create a neighbor list with 100-meter radius
# The distance is up to your discretion, you could find the average distance between points and use that as the radius


weights <- nb2listw(dists, style = "B") # points are on a binary, 1 if they are within your radius, 0 if not

moran.test(x = points$numeric_vector, listw=weights, zero.policy=T)
# x must be a numeric variable in the points data frame

```
